Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2024-04-22 paper/manuscript.qmd
Deleted unwanted reference to meta-analysis
abstract and format

title: "Is Phenomenological Control related to Illusion Sensitivity"

@@

abstract: "Blablablabla blablabla todo todo todo."

@@

keywords: [keyword1, keyword2, keyword3]

@@

masked-citations:
  - schneider2012cattell
  - schneider2015intelligence

@@

Such computational model(s) propose to conceptualize illusions as stimuli providing weak or conflicting sensory evidence [@sundareswara2008; @gershman2012] that bias perception toward prior knowledge. In other words, the weight of priors, in the form of perceptual knowledge about the world (e.g., internalized rules of perspective) is amplified when the sensory input is confusing. For instance, in the Müller-Lyer illusion, we “compute” the two (actually identical) lines as being of different lengths because the line flanked with converging fins is misinterpreted as being further away [@notredame2014]. In this context, measuring sensitivity to illusion can be operationalized as indexing the parameters of the Bayesian inference process (e.g., prior precision).

These accounts also provide a compelling framework to explain existing findings reporting interindividual variability in the sensitivity to illusions. Indeed, several studies suggest a relationship between psychopathology and lower illusion sensitivity. Schizophrenic patients, for instance, constantly exhibit lower scores on visual illusion tasks [@costa2023]. Proposed explanations, under this framework, emphasise alterations in metacognitive components whereby greater confidence is given to sensory evidence over prior knowledge, which show as reduced susceptibility to the effect of illusions [@notredame2014]. While evidence for lower visual sensitivity is less established in autistic individuals [@gori2016], reduced biases in the effects of illusions have also been attributed to diminished influence of top-down processes such as prior knowledge [@mitchell2010] and greater emphasis on sensory information [@palmer2017]. Evidence beyond psychopathology suggests illusion sensitivity variation in the general population, which could be correlated for instance with personality traits such as higher agreeableness and honest-humility [@makowski2023] and open-mindness and flexibility of cognitive control [@shoshina2014].

However, the exact nature of this interindividual variability and its potential origin remains unclear. The somewhat mixed evidence in the literature regarding its generalizability and strength could be related to the variety of the paradigms used and the type of processes being mobilised [@makowski2023]. Indeed, traditional methods frequently focus on participant’s experience by prompting them to assess the difference between two identical targets, estimate the target’s physical properties, or adjust the targets to match a reference stimulus [@todorovic2020]. Relying on metacognitive judgments about one’s subjective experiences adds an additional layer to the measure that might not be desired when attempting to measure illusion sensibility. Moreover, paradigms often face challenges in diversifying the illusory effects (i.e., using multiple stimuli to experimentally manipulate the strength of the illusion) and the illusion types (i.e., using various illusions, such as Müller-Lyer, Ebbinghaus, Delboeuf which might rely on a different admixture of mechanisms), hindering the potential of obtaining a comprehensive, valid, and reliable measure of illusion sensitivity [@makowski2023].

The “Illusion Game” paradigm [@makowski2023] has been recently developed to measure illusion sensitivity to various illusion types through its behavioural impact (on response time and error rate) in a perceptual decision task (where participants have to respond as fast as possible; e.g., “which of the left or right circles is bigger”). The stimuli for different classical illusions are created using the Pyllusion software [@makowski2021], which allows researchers to modulate the strength of the illusion as a continuous dimension, independently from the difficulty of the perceptual task. This paradigm, inspired by psychophysics, lends itself to the computational modelling of illusion sensitivity through its interference effect, hopefully bypassing some of the metacognitive processes at stake in other paradigms.

Interestingly, the fact that inter-individual variability in illusion sensitivity seems to persist in this task suggests that it is not solely explained by metacognitive abilities difference, and gives rise to the following question: is the variability in illusion sensitivity related to low-level perceptual processes (e.g., baseline precision of perceptual priors), or rather to the ability to actively control and “resist” the illusion in a top-town fashion in order to achieve the task at hand (higher-level modulation of the perceptual inference parameters). If the latter is true, then illusion sensitivity measured in contexts with strong task-demand characteristics, e.g., in paradigms where participants’ performance is explicitly or implicitly assessed (i.e., where there is an incentive to downplay the illusion effect) might correlate with one’s ability to alter one’s subjective experience following suggestions - a mechanism referred to as “phenomenological control”.

@@

To encourage the empirical exploration of our ability and tendency to alter our phenomenological experience and further accelerate investigations away from the hypnotic context, @lush2021 adapted the Sussex-Waterloo Scale of Hypnotisability [SWASH**,** @lush2018] by removing all its references to hypnosis, to measure trait phenomenological control. This newly developed phenomenological control scale (PCS) consists of 10 imaginative suggestions followed by subjective ratings for each suggestion on a 6-point Likert scale and has been demonstrated to be compatible with online experiments [@lush2022].

@@

# **Methods**

## **Participants**

@@

## **Procedure**

The experiment’s setup follows the born-open principle [@deleeuw2023]. The online experiment, implemented entirely using JsPsych [@de2015jspsych], has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. Participants’ raw data files are then automatically stored on a European server on OSF; and the deanonymized and ready-to-use data (downloaded from OSF and formatted using a reproducible script) will be made available in the same GitHub repository, alongside the scripts used for the analysis.

@@

**Phenomenological Control Scale (PCS)**. Participants will be asked to put on their headphones and await further auditory instructions. The PCS procedure starts with a recorded introduction explaining that a series of tests will be applied to evaluate how experiences can be created through imagination. This will be followed by 10 suggestions in a fixed order [see @lush2021], such as “now extend your arms ahead of you, with palms facing each other, hands about a foot apart” and “as you sit comfortably in your chair with your eyes closed, a picture of two balls will be displayed on the computer screen”. Once the 10 suggestions are completed, participants will be asked to rate their subjective experiences and response to each suggestion on a 6-points Likert scale. Phenomenological control will be indexed by averaging the scores from the 10 scales.

**Illusion Game**. The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one’s abilities.

For each illusion type, two continuous dimensions are orthogonally manipulated [see @makowski2021 for details on the rationale and execution], namely task difficulty and illusion strength, so that each trial corresponds to a unique combination. Task difficulty corresponds to the difficulty of the perceptual decision (e.g., if the task is to select the longest red line, task difficulty corresponds to how the lines are objectively different). Illusion strength corresponds to the degree to which the illusion elements (e.g., the black arrow lines in Müller-Lyer) are interfering with the aforementioned task. Note that the illusion effect can be “incongruent” (biasing perception in the direction of the incorrect response) or “congruent” (facilitating, i.e., biasing perception in the direction of the correct response). Participants respond with a key arrow (left vs. right; or up vs. down), and their reaction time (RT) and accuracy are recorded.

Visual illusion sensitivity will be measured as the average error rate in the incongruent condition, separately for the 3 illusion types. Although the error rate is arguably a crude score, which does not take into account the effect of varying illusion strength, the interaction with task difficulty and the possible adjustments in response strategy (speed-accuracy tradeoff), it is also the most simple and easy to reproduce, hence its usage as our primary outcome for the current preregistration.

@@

## **Data Analysis**

@@

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. If the RT is collapsed to the left (i.e., has \> ⅓ of ultra-fast responses - typically \< 200 ms) in the first block, the entire participant will be discarded (suggesting that they did not properly do the text), but if only the second block is bad, then only the second block will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision). In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @thériault2024].

After removing problematic participants and trials, the outcome measures (PC and VI sensitivity scores) will be computed and the Bayesian correlation (with medium prior on the coefficient, i.e., r-scale parameter set to 1/3) will be computed [using the BayesFactor package; @BayesFactor]**.** Following @lush2022, we expect to collect evidence against (BF10 \<= 1/3) a relationship between PCS and VI sensitivity. Data analysis will be carried out using R, using tidyverse [@tidyverse] and easystats [@easystats; @parameters; @correlation; @bayestestR; @datawizard]. The analysis script and additional information are available at https://github.com/RealityBending/IllusionGamePhenomenologicalControl.

>>>>

title: "Testing the Relationship between Phenomenological Control related to Illusion Sensitivity"

@@

abstract: "
Visual illusions highlight how easily our conscious experience can be altered with respect to perceptual reality. Despite sharing in-principle mechanisms with phenomenological control, i.e., the ability to alter our perceptual experience to match task demands or expectations, research tying the two remains scarce. This study aims to replicate and expand Lush et al. (2022) reporting an absence of correlation between phenomenological control (measured using the Phenomenological Control Scale) and illusion sensitivity to different illusion types. *[N participants were recruited in an online study. Results will added in the final version of the manuscript]*.
"

@@

keywords: [illusion sensitivity, visual illusions, phenomenological control, suggestibility, hypnotizability]

@@

@@

Such computational model(s) propose to conceptualize illusions as stimuli providing weak or conflicting sensory evidence [@sundareswara2008; @gershman2012] that bias perception toward prior knowledge. In other words, the weight of priors, in the form of perceptual knowledge about the world (e.g., internalized rules of perspective) is amplified when the sensory input is confusing. For instance, in the Müller-Lyer illusion, we "compute" the two (actually identical) lines as being of different lengths because the line flanked with converging fins is misinterpreted as being further away [@notredame2014]. In this context, measuring sensitivity to illusion can be operationalized as indexing the parameters of the Bayesian inference process (e.g., prior precision).

These accounts also provide a compelling framework to explain existing findings reporting interindividual variability in the sensitivity to illusions. Indeed, several studies suggest a potential link with psychopathology, in particular schizophrenia [@costa2023] and autism [@gori2016], in which the reported lower sensitivity to illusions has been attributed to a diminished influence of top-down processes such as prior knowledge [@mitchell2010] and a greater emphasis on (i.e., precision of) sensory information [@palmer2017]. Evidence beyond psychopathology also suggests variability in the general population, potentially correlated with personality traits such as agreeableness and honest-humility [@makowski2023], as well as cognitive abilities [@shoshina2014].

However, the exact nature of this interindividual variability and its potential origin remains unclear. The somewhat mixed evidence in the literature regarding its generalizability and strength could be related to the variety of the paradigms used and the type of processes being mobilised [@makowski2023]. Indeed, traditional methods frequently focus on participant's experience by prompting them to assess the difference between two identical targets, estimate the target's physical properties, or adjust the targets to match a reference stimulus [@todorovic2020]. Relying on metacognitive judgments about one's subjective experiences adds an additional layer to the measure that might not be desired when attempting to measure illusion sensibility. Moreover, paradigms often face challenges in diversifying the illusory effects (i.e., using multiple stimuli to experimentally manipulate the strength of the illusion) and the illusion types (i.e., using various illusions, such as Müller-Lyer, Ebbinghaus, Delboeuf which might rely on a different admixture of mechanisms), hindering the potential of obtaining a comprehensive, valid, and reliable measure of illusion sensitivity [@makowski2023].

The "Illusion Game" paradigm [@makowski2023] has been recently developed to measure illusion sensitivity to various illusion types through its behavioural impact (on response time and error rate) in a perceptual decision task (where participants have to respond as fast as possible; e.g., "which of the left or right circles is bigger"). The stimuli for different classical illusions are created using the *Pyllusion* software [@makowski2021], which allows researchers to modulate the strength of the illusion as a continuous dimension, independently from the difficulty of the perceptual task. This paradigm, inspired by psychophysics, lends itself to the computational modelling of illusion sensitivity through its interference effect, hopefully bypassing some of the metacognitive processes at stake in other paradigms.

Interestingly, the fact that inter-individual variability in illusion sensitivity seems to persist in this task suggests that it is not solely explained by metacognitive abilities difference, and gives rise to the following question: is the variability in illusion sensitivity related to low-level perceptual processes (e.g., baseline precision of perceptual priors), or rather to the ability to actively control and "resist" the illusion in a top-town fashion in order to achieve the task at hand (higher-level modulation of the perceptual inference parameters). If the latter is true, then illusion sensitivity measured in contexts with strong task-demand characteristics, e.g., in paradigms where participants' performance is explicitly or implicitly assessed (i.e., where there is an incentive to downplay the illusion effect) might correlate with one's ability to alter one's subjective experience following suggestions - a mechanism referred to as "phenomenological control".

@@

To encourage the empirical exploration of our ability and tendency to alter our phenomenological experience and further accelerate investigations away from the hypnotic context, @lush2021 adapted the Sussex-Waterloo Scale of Hypnotisability [SWASH, @lush2018] by removing all its references to hypnosis, to measure trait phenomenological control. This newly developed phenomenological control scale (PCS) consists of 10 imaginative suggestions followed by subjective ratings for each suggestion on a 6-point Likert scale and has been demonstrated to be compatible with online experiments [@lush2022].

@@

# Methods

## Participants

@@

## Procedure

The experiment's setup follows the born-open principle [@deleeuw2023]. The online experiment, implemented entirely using *JsPsych* [@de2015jspsych], has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. Participants' raw data files are then automatically stored on a European server on OSF; and the deanonymized and ready-to-use data (downloaded from OSF and formatted using a reproducible script) will be made available in the same GitHub repository, alongside the scripts used for the analysis.

@@

### Phenomenological Control Scale (PCS)

Participants will be asked to put on their headphones and await further auditory instructions. The PCS procedure starts with a recorded introduction explaining that a series of tests will be applied to evaluate how experiences can be created through imagination. This will be followed by 10 suggestions in a fixed order [see @lush2021], such as "now extend your arms ahead of you, with palms facing each other, hands about a foot apart" and "as you sit comfortably in your chair with your eyes closed, a picture of two balls will be displayed on the computer screen". Once the 10 suggestions are completed, participants will be asked to rate their subjective experiences and response to each suggestion on a 6-points Likert scale. Phenomenological control will be indexed by averaging the scores from the 10 scales.

### Illusion Game

The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one's abilities.

For each illusion type, two continuous dimensions are orthogonally manipulated [see @makowski2021 for details on the rationale and execution], namely task difficulty and illusion strength, so that each trial corresponds to a unique combination. Task difficulty corresponds to the difficulty of the perceptual decision (e.g., if the task is to select the longest red line, task difficulty corresponds to how the lines are objectively different). Illusion strength corresponds to the degree to which the illusion elements (e.g., the black arrow lines in Müller-Lyer) are interfering with the aforementioned task. Note that the illusion effect can be "incongruent" (biasing perception in the direction of the incorrect response) or "congruent" (facilitating, i.e., biasing perception in the direction of the correct response). Participants respond with a key arrow (left vs. right; or up vs. down), and their reaction time (RT) and accuracy are recorded.

Visual illusion sensitivity will be measured as the average error rate in the incongruent condition, separately for the 3 illusion types. Although the error rate is arguably a crude score, which does not take into account the effect of varying illusion strength, the interaction with task difficulty and the possible adjustments in response strategy (speed-accuracy trade off), it is also the most simple and easy to reproduce, hence its usage as our primary outcome for the current preregistration.

@@

## Data Analysis

@@

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. If the RT is collapsed to the left (i.e., has \> 1/3 of ultra-fast responses - typically \< 200 ms) in the first block, the entire participant will be discarded (suggesting that they did not properly do the text), but if only the second block is bad, then only the second block will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision). In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @theriault2024].

After removing problematic participants and trials, the outcome measures (PC and VI sensitivity scores) will be computed and the Bayesian correlation (with medium prior on the coefficient, i.e., r-scale parameter set to 1/3) will be computed [using the *BayesFactor* package; @BayesFactor]. Following @lush2022, we expect to collect evidence against (BF10 \<= 1/3) a relationship between PCS and VI sensitivity. Data analysis will be carried out using R, using *tidyverse* [@tidyverse] and *easystats* [@easystats; @parameters; @correlation; @bayestestR; @datawizard]. The analysis script and additional information are available at *https://github.com/RealityBending/IllusionGamePhenomenologicalControl*.

\pagebreak
